Molecular dynamics simulation of synchronization in driven particles

### 2020-12-29 manuscript/manuscript_ajp_2020.tex
added schematic of simulation!

%-----------------------------------------------------------------
%transition from synchronized to chaotic
%describe types of physical systems - particle types
%%In this work,
%we perform 
%numerical simulations of confined, driven particles
%to model a variety of physical phenomena.
%-----------------------------------------------------------------------
%paper outline
%In the following paper
We describe
our molecular dynamics model in Section~\ref{sec:MD}
In this section, we restrict our discussion to a single particle,
and include additional information
for multiple particles in Section~\ref{sec:sync}.
%including various  and confining environments.
We summarize
our results 
in Section~\ref{sec:results} 
including synchronized motion of a single confined particle
driven across a periodic landscape in 
Section~\ref{sec:one} and
multiple interacting particles
in Section~\ref{sec:sync},
including stationary propagation of high density kinks
in Sec.~\ref{sec:kink}.
We include code to simulate
and visualize the landscape and dynamics in this section
and supplementary material.
%we
%demonstrate how uniform environments and applied forces
%and create synchronized flow patterns.
We present these results using standard tools of non-linear oscillators
such phase diagrams of velocity versus position.
%PHASE DIAGRAMS SHOWING COUPLING, OR QUASI-PERIODIC COUPLING
%INDICATE COOPERATIVE MOTION IN THE Q1D SUBSTRATE
%In Section~\ref{sec:quasiperiod},
%we show how an aperiodic landscape modifies the particle dynamics.
%Finally we explore the transition to chaotic dynamics in 
%Section~\ref{sec:chaos}
%and conclude 
%in Section~\ref{sec:conclusion}. 
In each section
we suggest exercises for interested students,
as summarized in Section~\ref{sec:problems}.

@@

\section{Molecular Dynamics Simulation}
\label{sec:MD}
In this section we describe the dynamics of the particles
using the model forces
and the motion using an modified form of Newton's second law
suitable for microscopic particles in a suspending fluid.

We use a classical two-dimensional model for 
studying the dynamics of $N$ interacting particles.
Particles are confined in a two-dimensional (2D) 
simulation of area $A = L \times L$ where $L=36.5 a_0$
where $a_0$ is a dimensionless unit of length.
An individual particle $i$ has
position $\vec{r}_i = x_i \hat{x} + y_i \hat{y}$
and velocity $\vec{v}_i = d\vec{r}_i/dt$.  
Particles are subject to
periodic boundary conditions
such that a particle leaving the edge of the system is mapped
back to a position within the simulation 
by the transformation $x_i+L \rightarrow x_i$ and $y_i+L \rightarrow y_i$.
%and particles interact across PBC.

@@

We confine the particles using a position dependent 
potential energy function, called a landscape or substrate.
The confining force %arises from a potential function 
 %creating a landscape of potential minima and maxima
%that modify the force on a particle as a function of position
is calculated as 
 $\vec{F}^{l}_i(\vec{r}_i) = -\nabla U_l(\vec{r}_i)$.
 %The landscape potential energy $U(\vec{r})$ is static 
 %with fixed minima and maxima
 %that are periodic or quasi-periodic,
 %as described in Sec.~\ref{sec:results}.
% 
The periodic landscape is modulated in the y-direction as
%% %that used in Sec.~\ref{sec:one}
% where 
 \begin{equation}
   U_{washboard}(y) = U_0 \cos{(N_p \pi y / L)}
     \label{eq:ysubstrate}
\end{equation}
 where $N_p$ are the number of periods,
 % $L$ is the period length,
 and $U_0$ is an adjustable parameter
 to set the depth of the troughs.
 We define the substrate strength to be
 the force 
 $A_p = 2\pi U_0/L$.
 %

%because 
We model the 
particle dynamics 
with an overdamped
equation of motion.
%integrated with the Verlet method.
%The suspending fluid is highly viscous
%and exerts a damping force on the particles equivalent
%to a linear drag force.
%Overdamped dynamics is a common assumption in models of colloidal particles.
%This is an assumption that
The term ``overdamped'' or critically damped indicates
that an oscillator cannot undergo periodic
undulations due to some external force. %such as
%suspension in a viscous fluid \cite{}.
In our colloidal model we assume
the damping comes from the viscosity of the fluid,
where a sphere of radius $0.5 \mu m$ has a Reynolds number
of $R \approx 10^{-6}$, indicating that linear drag dominates its motion.
See Section~\ref{sec:problems} Exercise~\ref{ex:reynolds} for more information
about the Reynold's number.
We model this
nonconservative force,
with the linear friction $\vec{F}_{drag} = -\eta \vec{v}_i$
sufficient so that
the acceleration of the particle is zero.
Such a model is appropriate when the particles
are small and the viscosity of the suspending medium is high.
% [finish... discussion of Reynolds number]
This model should be familiar to readers
modeling Millikan's oil drops
in a standard classical mechanics text \cite{Taylor2005}.
%This model is encountered in standard texts,
%for instance Ch. 2 of Taylor's Classical Mechanics.
%more precise: 6\pi \eta a \vec{v}

@@

\begin{table}[h!]
\centering
\caption{Simulation units.}
\begin{ruledtabular}
\begin{tabular}{c c p{5cm}}
Quantity & Unit \\
\hline
length &  $a_0$ \\
%electric potential & $V(r_{ij}) = E_0/r_{ij}$ \\
energy & $E_0 = q^2{Z^*}^2/4\pi \epsilon \epsilon_0 a_0$ \\
dimensionless interaction strength & $q$ \\
effective colloidal charge & $Z^*$ \\
solvent dielectric constant & $\epsilon \epsilon_0$\\
force & $E_0 / a_0$\\
viscosity/damping constant & $\eta$ \\
time &  $\eta / E_0$ \\
velocity &  $ E_0 / \eta a_0$ \\
\end{tabular}
\end{ruledtabular}
\label{tab:1}
\end{table}

>>>>

%-----------------------------------------------------------------
%transition from synchronized to chaotic
%describe types of physical systems - particle types
%%In this work,
%we perform 
%numerical simulations of confined, driven particles
%to model a variety of physical phenomena.
%-----------------------------------------------------------------------
%paper outline
%In the following paper
We describe
our molecular dynamics model for a single particle in Section~\ref{sec:MD}.
%In this section, we restrict our discussion to a single particle,
%and include additional information
%for multiple particles in Section~\ref{sec:sync}.
%including various  and confining environments.
We summarize
our results 
in Section~\ref{sec:results} 
including synchronized motion of a single confined particle
driven across a periodic landscape in 
Section~\ref{sec:one}.
In Section~\ref{sec:sync}
we 
extend the model to include multiple particles,
and show how confined 
interacting particles respond to the same applied driving force.
%including stationary propagation of high density kinks
%in Sec.~\ref{sec:kink}.
%We include code to simulate
%and visualize the landscape and dynamics in this section
%and supplementary material.
%we
%demonstrate how uniform environments and applied forces
%and create synchronized flow patterns.
%We present these results using standard tools of non-linear oscillators
%such phase diagrams of velocity versus position.
%PHASE DIAGRAMS SHOWING COUPLING, OR QUASI-PERIODIC COUPLING
%INDICATE COOPERATIVE MOTION IN THE Q1D SUBSTRATE
%In Section~\ref{sec:quasiperiod},
%we show how an aperiodic landscape modifies the particle dynamics.
%Finally we explore the transition to chaotic dynamics in 
%Section~\ref{sec:chaos}
%and conclude 
%in Section~\ref{sec:conclusion}. 
We conclude with
exercises for interested students 
in Section~\ref{sec:problems}.

@@

\section{Molecular Dynamics Simulation}
\label{sec:MD}

\begin{center}
\begin{figure}[h!]
\centering
\includegraphics[width=\columnwidth]{landscape.pdf}
\caption{
  Schematic of the simulation
  of a single particle
  driven across a washboard potential
  energy landscape.
  (a) View of the $x-y$ plane. %of the simulation
  The time-dependent applied driving force $\vec{F}^D$
  is parallel to the y-axis.
  The force due to the landscape %$\vec{F}^l$
  is calculated due to the gradient of the potential
  energy 
  $\vec{F}^l = -\nabla U(\vec{r})$.
  (b) The potential energy function
  along the y-axis $U_l(y)$
  with period $\lambda = L/N_p$
  where the number of periods is $N_p = 3$.
  }
\label{fig:landscape0}
\end{figure}
\end{center}

We use a classical two-dimensional model for 
studying the dynamics of $N$ interacting particles,
using the net force on each particle to calculate
its trajectory.
Particles are confined in a two-dimensional (2D) 
simulation of area $A = L \times L$ where $L=36.5 a_0$
where $a_0$ is a dimensionless unit of length.
An individual particle $i$ has
position $\vec{r}_i = x_i \hat{x} + y_i \hat{y}$
and velocity $\vec{v}_i = d\vec{r}_i/dt$.
The edges of the system are treated with
periodic boundary conditions
such that a particle leaving the edge of the system is mapped
back to a position within the simulation 
by the transformation $x_i+L \rightarrow x_i$ and $y_i+L \rightarrow y_i$.
%and particles interact across PBC.
We show a schematic of the system in Fig.~\ref{fig:0}

We confine the particles using a position dependent 
potential energy function, called a landscape or substrate.
The confining force %arises from a potential function 
 %creating a landscape of potential minima and maxima
%that modify the force on a particle as a function of position
is calculated as 
 $\vec{F}^{l}_i(\vec{r}_i) = -\nabla U_l(\vec{r}_i)$.
 %The landscape potential energy $U(\vec{r})$ is static 
 %with fixed minima and maxima
 %that are periodic or quasi-periodic,
 %as described in Sec.~\ref{sec:results}.
% 
The periodic landscape is modulated in the y-direction as
%% %that used in Sec.~\ref{sec:one}
% where 
 \begin{equation}
   U(y) = U_0 \cos{(N_p \pi y / L)}
     \label{eq:ysubstrate}
\end{equation}
 where $N_p$ are the number of periods,
 % $L$ is the period length,
 and $U_0$ is an adjustable parameter
 to set the depth of the troughs.
 We define the substrate strength to be
 the maximum force the substrate can exert on a particle 
 $A_p = 2\pi U_0/L$.
 %

@@

%because 
We model the 
particle dynamics 
with an overdamped
equation of motion.
%integrated with the Verlet method.
%The suspending fluid is highly viscous
%and exerts a damping force on the particles equivalent
%to a linear drag force.
%Overdamped dynamics is a common assumption in models of colloidal particles.
%This is an assumption that
The term `overdamped or critically damped indicates
that an oscillator cannot undergo periodic
undulations due to some external force. %such as
%suspension in a viscous fluid \cite{}.
In our colloidal model we assume
the damping comes from the viscosity of the fluid,
where a sphere of radius $0.5 \mu m$ has a Reynolds number
of $R \approx 10^{-6}$, indicating that linear drag dominates its motion.
See Section~\ref{sec:problems} Exercise~\ref{ex:reynolds} for more information
about the Reynold's number.
We model this
nonconservative force,
with the linear friction $\vec{F}_{drag} = -\eta \vec{v}_i$
sufficient so that
the acceleration of the particle is zero.
Such a model is appropriate when the particles
are small and the viscosity of the suspending medium is high.
% [finish... discussion of Reynolds number]
This model should be familiar to readers
modeling Millikan's oil drops
in a standard classical mechanics text \cite{Taylor2005}.
%This model is encountered in standard texts,
%for instance Ch. 2 of Taylor's Classical Mechanics.
%more precise: 6\pi \eta a \vec{v}

@@

\begin{table}[h!]
\centering
\caption{Simulation units.}
\begin{ruledtabular}
\begin{tabular}{c c p{5cm}}
Quantity & Unit \\
\hline
length &  $a_0$ \\
energy & $E_0$ \\
%electric potential & $V(r_{ij}) = E_0/r_{ij}$ \\
%energy & $E_0 = q^2{Z^*}^2/4\pi \epsilon \epsilon_0 a_0$ \\
%dimensionless interaction strength & $q$ \\
%effective colloidal charge & $Z^*$ \\
%solvent dielectric constant & $\epsilon \epsilon_0$\\
force & $E_0 / a_0$\\
viscosity/damping constant & $\eta$ \\
time &  $\eta / E_0$ \\
velocity &  $ (E_0 / \eta) a_0$ \\
\end{tabular}
\end{ruledtabular}
\label{tab:1}
\end{table}
